---
title: "Low Dimensional Projection of Cytometry Samples"
author:
    - name: Philippe Hauchamps
    - name: Laurent Gatto
package: CytoPipeline
abstract: >
 This vignette describes the functionality implemented in the `CytoMDS`
 package. `CytoMDS` provides support for low dimensional projection of a set  
 of cytometry samples, using concepts such as Earth Mover's (EMD) distance,   
 and Multi Dimensional Scaling (MDS).  
 This vignette is distributed under a CC BY-SA 4.0 license.
output:
  BiocStyle::html_document:
    toc_float: true
bibliography: CytoMDS.bib
vignette: >
  %\VignetteIndexEntry{Low Dimensional Projection of Cytometry Samples}
  %\VignetteEngine{knitr::rmarkdown}
  %%\VignetteKeywords{FlowCytometry, QualityControl, DimensionReduction, multidimensionalScaling, Software, Visualization}
  %\VignetteEncoding{UTF-8}
---

```{r style, echo = FALSE, results = 'asis'}
BiocStyle::markdown()
```
# Installation

To install this package, start R and enter (uncommented):

```{r}
# if (!require("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# 
# BiocManager::install("CytoMDS")
```

Loading dependent packages for the present vignette...

```{r libs}
library(CytoPipeline)
library(CytoMDS)
```

# Introduction

TO DO



# Simulated datasets

We now simulate 2 datasets by combining events from the two Donor samples of the
OMIP021 original dataset.

```{r OMIP021Trans}
data(OMIP021Samples)

#outputDir <- base::tempdir()

transList <- estimateScaleTransforms(
    ff = OMIP021Samples[[1]],
    fluoMethod = "estimateLogicle",
    scatterMethod = "linearQuantile",
    scatterRefMarker = "BV785 - CD3")

OMIP021Trans <- CytoPipeline::applyScaleTransforms(
    OMIP021Samples, 
    transList)
```


```{r OMIPSim1}
nSample <- 10
ffList <- list()
for (i in seq_len(nSample)) {
    ffList[[i]] <- CytoPipeline::subsample(
                OMIP021Trans[[(i+1)%%2+1]],
                nEvents = 1000,
                seed = i)[,1:22]
}
OMIP021Sim1 <- as(ffList, "flowSet")

pData1 <- data.frame(name = paste0("S", seq_len(nSample)),
                     clust = rep(c(1,2), 5))
```

```{r OMIPSim2}
nSample <- 11
ffList <- list()
for (i in seq_len(nSample)) {
    if (i==1) {
        ffList[[i]] <- OMIP021Trans[[1]]
    } else if (i==nSample) {
        ffList[[i]] <- OMIP021Trans[[2]]
    } else {
        ff1 <- CytoPipeline::subsample(
            OMIP021Trans[[1]],
            nEvents = 1000 * (nSample-i)/(nSample-1),
            seed = i)[,1:22]
        ff2 <- CytoPipeline::subsample(
            OMIP021Trans[[2]],
            nEvents = 1000 * (i-1)/(nSample-1)
        )
        
        ffList[[i]] <- CytoPipeline::aggregateAndSample(
            flowCore::flowSet(ff1, ff2),
            nTotalEvents = 1000000 # big number to have a simple aggregation
        )[,1:22]
    }
}
OMIP021Sim2 <- as(ffList, "flowSet")

pData2 <- data.frame(name = paste0("S", seq_len(nSample)),
                     origin = c("Raw", rep("Sim", nSample-2), "Raw"),
                     rawLabel = c("D1", rep("", nSample-2), "D2"))

```

# calculating distances between samples

```{r distCalc}
pwDist1 <- pairwiseEMDDist(
    OMIP021Sim1, verbose = TRUE)
```

```{r distShow}
round(pwDist1, 2)
```

# calculating MDS projection

```{r MDSCalc}

```






# Session information {-}

```{r sessioninfo, echo=FALSE}
sessionInfo()
```


# References {-}
