---
title: "Low Dimensional Projection of Cytometry Samples"
author:
    - name: Philippe Hauchamps
    - name: Laurent Gatto
package: CytoMDS
abstract: >
 This vignette describes the functionality implemented in the `CytoMDS`
 package. `CytoMDS` provides support for low dimensional projection of a set  
 of cytometry samples, using concepts such as Earth Mover's (EMD) distance,   
 and Multi Dimensional Scaling (MDS).  
 This vignette is distributed under a CC BY-SA 4.0 license.
output:
  BiocStyle::html_document:
    toc_float: true
bibliography: CytoMDS.bib
vignette: >
  %\VignetteIndexEntry{Low Dimensional Projection of Cytometry Samples}
  %\VignetteEngine{knitr::rmarkdown}
  %%\VignetteKeywords{FlowCytometry, QualityControl, DimensionReduction, multidimensionalScaling, Software, Visualization}
  %\VignetteEncoding{UTF-8}
---

```{r style, echo = FALSE, results = 'hide'}
BiocStyle::markdown()
```
# Installation

To install this package, start R and enter (un-commented):

```{r}
# if (!require("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# 
# BiocManager::install("CytoMDS")
```

Loading dependent packages for the present vignette...

```{r libs, results = FALSE}
library(CytoPipeline)
library(CytoMDS)
library(ggplot2)
```

# Introduction

The `CytoMDS` package  implements a low dimensional visualization of a set
of cytometry samples, in order to visually assess the 'distances' between them.
This, in turn, can greatly help the user to identify quality issues 
like batch effects or outlier samples, and/or check the presence of potential 
sample clusters that might align with the experimental design.  

The `CytoMDS` algorithm combines, on the one hand, the concept of 
*Earth Mover's Distance (EMD)* [@Orlova2016-jm], a.k.a. *Wasserstein metric* and, 
on the other hand, the metric *Multi Dimensional Scaling (MDS)* algorithm 
for the low dimensional projection [@De_Leeuw2009-aw]. 

Also, the package provides some diagnostic tools for both checking the quality 
of the MDS projection, as well as tools to help with the interpretation of 
the axes of the projection (see below sections).

# Illustrative datasets

Illustrative data sets that will be used throughout this vignette are derived 
from a reference public dataset accompanying the OMIP-021 
(Optimized Multicolor Immunofluorescence Panel 021) article [@Gherardin2014-pj].  

A sub-sample of this public dataset is built-in in the `CytoPipeline` package 
[@CytoPipeline] as the `OMIP021` dataset.

In the `CytoMDS`package, as in the current vignette, 
matrices of flow cytometry events intensities are stored as 
`flowCore::flowFrame` [@flowCore] objects. 

Note that the `OMIP021` dataset only contains two samples, 
from two human donors. Therefore, in order to be able to meaningfully illustrate 
 `CytoMDS` use cases, we will be building data sets with more samples, 
 simulated by combining events, sampled from the two original 
 `OMIP021` samples.  

The first step consists in scale transforming the two raw flow frames, using 
estimated scale transformations. Indeed, distances between samples make more 
sense with scaled transformed signal, in which distributional differences are 
much more obvious. In order to transform the signal of the different channels, 
here we use the `estimateScaleTransforms()` function from the `CytoPipeline` 
package. However, this can be done using any standard package for flow 
cytometry data.

```{r OMIP021Trans}
data(OMIP021Samples)

#outputDir <- base::tempdir()

transList <- estimateScaleTransforms(
    ff = OMIP021Samples[[1]],
    fluoMethod = "estimateLogicle",
    scatterMethod = "linearQuantile",
    scatterRefMarker = "BV785 - CD3")

OMIP021Trans <- CytoPipeline::applyScaleTransforms(
    OMIP021Samples[,c(1:16,18,20:22)], # removing 'EMPTY' channels
    transList)
```

We now create two simulated data sets, of 20 samples each, by combining events 
from the two samples of the OMIP021 original data set. 

We also attached to each sample, some 'phenoData' dataframe, 
describing each sample with some chosen variables.

The first simulated data set, which is called here `OMIP021Sim1`, 
is built by sub-sampling 5 times from each of the two samples (alternatively). 

```{r OMIPSim1}
nSample <- 20
ffList <- list()

# to allow for reproducibility of sample()
set.seed(0)

for (i in seq_len(nSample)) {
    ffList[[i]] <- CytoPipeline::subsample(
                OMIP021Trans[[(i+1)%%2+1]],
                nEvents = 1000,
                seed = i)
}
OMIP021Sim1 <- as(ffList, "flowSet")

pData1 <- data.frame(name = paste0("S", seq_len(nSample)),
                     original_sample = factor(rep(c(1,2), 5)))
```

The second one, `OMIP021Sim2` is built by mixing events from sample 1 
and sample 2 with respective proportions varying from 1 to 0 (resp. 0 to 1).

```{r OMIPSim2}
nSample <- 20
ffList <- list()
S1prop <- rep(0., nSample)

# to allow for reproducibility of sample()
set.seed(0)

for (i in seq_len(nSample)) {
    if (i==1) {
        ffList[[i]] <- OMIP021Trans[[1]]
        S1prop[i] <- 1
    } else if (i==nSample) {
        ffList[[i]] <- OMIP021Trans[[2]]
        S1prop[i] <- 0
    } else {
        ff1 <- CytoPipeline::subsample(
            OMIP021Trans[[1]],
            nEvents = 1000 * (nSample-i)/(nSample-1),
            seed = i)
        ff2 <- CytoPipeline::subsample(
            OMIP021Trans[[2]],
            nEvents = 1000 * (i-1)/(nSample-1),
            seed = i)
        S1prop[i] <- (nSample-i)/(nSample-1)
        
        ffList[[i]] <- CytoPipeline::aggregateAndSample(
            flowCore::flowSet(ff1, ff2),
            nTotalEvents = 1000000 # big number to have a simple aggregation
        )[,1:20]
    }
}
OMIP021Sim2 <- as(ffList, "flowSet")

pData2 <- data.frame(name = paste0("S", seq_len(nSample)),
                     origin = factor(c("Raw", rep("Sim", nSample-2), "Raw")),
                     rawLabel = c("D1", rep("", nSample-2), "D2"),
                     S1prop = S1prop)

```

# Calculating distances between samples

We can now calculate pairwise *Earth Mover's Distances (EMD)* 
between all samples of our simulated data sets.   

This is done by calling the `pairwiseEMDDist()` function.  
The latter function takes here a `flowCore::flowSet` as input parameter, 
which is a collection of all samples as `flowCore::flowFrame` objects. 
Note that, for heavy data sets that contains a lot of samples, this can create 
memory issues. In that case, there are other ways to call the 
`pairwiseEMDDist()` function (see 'Handling heavy data sets' section).

Using the `channels` argument, it is possible to restrict the EMD calculation 
to some of the channels. However, by default, all *signal* channels 
will be incorporated. *Signal* channels means all scatter and fluorescent 
channels. The known non-relevant channels for multivariate distribution 
distances, like time and other usual house-keeping channels, are automatically 
excluded from the calculation.

```{r distCalc}
pwDist1 <- pairwiseEMDDist(
    OMIP021Sim1, channels = NULL)

pwDist2 <- pairwiseEMDDist(
    OMIP021Sim2, channels = NULL)
```

The calculated distance is a symmetric square matrix, with as many rows 
(columns) as input samples 
(extract shown here below for `OMIP021Sim1` simulated data set).

```{r dist1Show}
round(pwDist1[1:10, 1:10], 2)
```

One relevant way to visualize this distance matrix is to draw the histogram 
of pairwise distances, as shown in the below plot, for the `OMIP021Sim1` 
simulated data set. Notice here the bi-modal distribution of the distances, due 
to the way the data set was generated (half of the samples originate from 
sub-sampling the first original sample, and the other half originates from 
sub-sampling of the second original sample.)

```{r dist1ShowHist, fig.align='center', fig.wide = TRUE}
distVec1 <- pwDist1[upper.tri(pwDist1)]
distVecDF1 <- data.frame(dist = distVec1)
pHist1 <- ggplot(distVecDF1, mapping = aes(x=dist)) + 
    geom_histogram(fill = "darkgrey", col = "black", bins = 15) + 
    theme_bw() + ggtitle("EMD distances for data set 1")
pHist1
```

The same type of pairwise distance histogram, but this time for the 
`OMIP021Sim12` simulated data set, looks much different: here the distribution 
is unimodal and more regular, as one can expect from the way this second 
simulated data set was generated.

```{r dist12ShowHist, fig.align='center', fig.wide = TRUE}
distVec2 <- pwDist2[upper.tri(pwDist2)]
distVecDF2 <- data.frame(dist = distVec2)
ggplot(distVecDF2, mapping = aes(x=dist)) + 
    geom_histogram(bins = 15, fill = "darkgrey", col = "black") + 
    theme_bw() + ggtitle("EMD distances for data set 2")
```

# Metric Multidimensional scaling

## Calculating the MDS projection

Once the pairwise distance matrix has been calculated, computing the *Multi 
Dimensional Scaling (MDS)* projection is done 
by calling the `computMetricMDS()` function. In its simplest form, only 
the distance matrix needs to be passed to the function. 
In that case, the number of dimensions to use in the MDS is automatically 
set in order to reach a specific value for a projection quality indicator, 
i.e. a *target pseudo R square*, which in turn is set by default set to 0.95 
(see Quality of projection - diagnostic tools section).  

Note that the *Smacof* algorithm [@De_Leeuw2009-aw], 
used to compute the MDS projection, is stochastic, so it is sensitive 
to the 'seed' used. Therefore, in cases where reproducible results from 
one run to another is required , it is advised to set the seed argument 
to a specific value.

```{r MDSCalc1}
mdsObj1 <- CytoMDS::computeMetricMDS(pwDist1, seed = 0)
```

## Plotting the MDS projection

Plotting the obtained MDS projection is done using `ggplotSampleMDS()`. 
If no phenoData is used, then, by default, numbers are used as labels, and 
the samples are represented as black dots.

```{r plotMDS1_simplest, fig.align='center', fig.asp = 1, fig.wide = TRUE}
ggplotSampleMDS(mdsObj1)
```

However, by providing a 'phenoData' dataframe to the `ggplotSampleMDS()` 
function, the corresponding variable can be used for highlighting sample points 
with different colours and/or shapes. Here below, the previous plot is enhanced 
with red and blue colours, dot and triangle shapes, distinguishing samples 
based on the value of the `original_sample` variable. Also, we have 
here added explicit labels to each data point, using the corresponding value 
of the `name` variable for each sample.


```{r plotMDS1_colors_shapes, fig.align='center', fig.asp = 0.8, fig.wide = TRUE}
ggplotSampleMDS(mdsObj1, 
                pData = pData1, 
                pDataForColour = "original_sample",
                pDataForShape = "original_sample",
                pDataForLabel = "name")
```

## Quality of projection - diagnostic tools

In order to be able to trust the projected distances obtained on the `CytoMDS` 
plots, a couple of projection quality indicators need to be taken into account: 
- the *pseudo RSquare* indicator shows what percentage of the variability 
contained in the pairwise distance matrix is actually shown in the projection. 
It is analog to the statistical *RSquare* for a linear regression model: 
the closer to one the *pseudo RSquare* is, the better.  

Note that the latter refers to the variability contained in __ALL__ 
dimensions of the MDS projection, not only the two plotted axes.
- *nDim* is the number of dimensions of the projection that was needed 
to obtain the corresponding *pseudo RSquare*
- the percentage of variation that is captured along each axis (coordinates), 
is to be interpreted with respect to the total variability __that is 
captured by the MDS projection__, not the total variability. 
For example, in the plot above, using 2 dimensions, the MDS projection is able 
to capture 97.01% (*pseudo RSquare*) of the initial variability contained 
in the calculated pairwise distance matrix. __Of these 97.01%__, 
90.96% is in turn captured by axis 1, and 9.04% is captured by axis 2.  

Another useful projection quality diagnostic tool is provided by the 
*Shepard diagram*. On this plot, each dot represents one distance 
between a sample pair, with as x coordinate the original (high dimensional) 
distance between the two samples, and as y coordinate the projected 
low dimensional distance between these two samples, as obtained by the MDS 
projection algorithm. In the *Shepard* diagram, an ideal situation corresponds 
to all points being located on the straight line passing through through 
the (0,0) and (1,1) points.  

```{r plotMDSShepard, fig.align='center', fig.wide = TRUE, fig.asp = 1}
ggplotSampleMDSShepard(mdsObj1)
```


## Additional options 

In order to show some of the additional options available to the user of 
`computeMetricMDS()` and `ggplotSampleMDS`(), let us project the pairwise 
sample distances of the `OMIP021Sim2` data set.

```{r MDSCalc2.noOption, fig.align='center', fig.asp = 0.85, fig.wide = TRUE}
mdsObj2 <- CytoMDS::computeMetricMDS(pwDist2)
ggplotSampleMDS(mdsObj2,
                pData = pData2, 
                pDataForLabel = "rawLabel",
                pDataForShape = "origin",
                pDataForColour = "S1prop")
```

In the subtitle of this latter plot, it is mentioned that the obtained R square 
of 95.41% was obtained thanks when using 3 dimensions 
in the Multi Dimensional Scaling. Therefore, one can visualize the MDS 
projection using any combination of two axes, 
for example axes 2 and 3, as below:

```{r MDSCalc2.axes3.4, fig.align='center', fig.asp = 0.85, fig.wide = TRUE}
ggplotSampleMDS(mdsObj2,
                projectionAxes = c(2, 3),
                pData = pData2, 
                pDataForLabel = "rawLabel",
                pDataForShape = "origin",
                pDataForColour = "S1prop")
```

Now it is also possible to impose the number of dimensions used in the MDS 
projection explicitly, for example to 2, as is shown below:

```{r MDSCalc2.nDim2, fig.align='center', fig.asp = 0.85, fig.wide = TRUE}
mdsObj2_2 <- CytoMDS::computeMetricMDS(pwDist2, nDim = 2)
ggplotSampleMDS(mdsObj2_2,
                pData = pData2, 
                pDataForLabel = "rawLabel",
                pDataForShape = "origin",
                pDataForColour = "S1prop")
```

Note that the obtained projection on 2 axes, although similar, is not exactly 
the same as the one obtained when visualizing the first two axis of the 
MDS projected before, on 3 dimensions. Actually, this is a feature of the 
Metric MDS projection, although it might appear a bit counter-intuitive 
at first.  

Finally, it is also possible to adjust the number of dimensions indirectly, 
by setting an explicit *pseudo Rsquare* target. In that case the algorithm will
increase the number of dimensions until reaching the required quality target.
The below example shows how to obtain a *pseudo R Square* of at least 0.99. 
Here the obtained number of dimensions is 6, instead of 3.

```{r MDSCalc2.Rsq99, fig.align='center', fig.asp = 0.85, fig.wide = TRUE}
mdsObj2_3 <- CytoMDS::computeMetricMDS(pwDist2, targetPseudoRSq = 0.99)
ggplotSampleMDS(mdsObj2_3,
                pData = pData2, 
                pDataForLabel = "rawLabel",
                pDataForShape = "origin",
                pDataForColour = "S1prop")
```

The corresponding Shepard diagram is obtained as below:

```{r biplot median, fig.align='center', fig.wide = TRUE, fig.asp = 1}
ggplotSampleMDSShepard(mdsObj2_3)
```



## Aid to interpreting projection axes

With MDS projections, it is possible to (try to) associate some axis directions 
to specific characteristics of the samples. The idea is to calculate the 
correlation of well chosen sample statistics w.r.t. the axes of projection, so 
that these correlations can be represented on a correlation circle, which is in 
turn overlaid on the projection plot. This plot set-up is called a 'bi-plot'.

In order to leverage on this functionality, the user first needs to calculate 
some statistics of interest for which they want to assess the association 
with the axis directions. Typically, one chooses channel specific statistics, 
like e.g. the mean, the standard deviation, or any quantile that might be 
of interest. However, any statistics that can be calculated for each sample 
can be used (number of events,...)

Here below, we provide an example where the user overlays the median of 
the different channels, on a bi-plot for the MDS projection obtained for
the first data set.  

On the bi-plot, each arrow - here representing a channel median - is located 
at coordinates equal to its Pearson correlation with the respective axis. 

Here, one can identify that the x axis has a strong positive correlation with 
the median of markers 'Viability', 'gdTCR', 'TCR Va7', 'CD45-RA', 'CD8a' and 
'CD27', and a strong negative correlation with the median of channels 'FSC-A', 
'SSC-A' and marker 'CD161'. The y axis has a strong negative correlation with 
the medians of markers 'CD28', and 'CD3'.

```{r plotMDS1_medians}
medians <- channelSummaryStats(OMIP021Sim1, statFUNs = median)
```

```{r plotMDS1_biplot, fig.align='center', fig.asp = 0.8, fig.wide = TRUE}
ggplotSampleMDS(mdsObj1, 
                pData = pData1, 
                pDataForColour = "original_sample",
                pDataForShape = "original_sample",
                displayPointLabels = FALSE,
                displayArrowLabels = TRUE,
                repelArrowLabels = TRUE,
                biplot = TRUE,
                extVariables = medians)
```

Note that, on the bi-plots, only the arrows of length greater or equal to 
a specific threshold (by default set at 0.8) are represented, in order to not 
overwhelm the plot with arrows, especially when the data sets contains a high 
dimensional panels.  

It is however possible to adjust this threshold by explicitly setting 
the `arrowThreshold` argument. For example, in the below plot, this threshold 
is set set to 0.9:

```{r plotMDS1_biplot_0.9, fig.align='center', fig.asp = 0.8, fig.wide = TRUE}
ggplotSampleMDS(mdsObj1, 
                pData = pData1, 
                pDataForColour = "original_sample",
                pDataForShape = "original_sample",
                displayPointLabels = FALSE,
                displayArrowLabels = TRUE,
                repelArrowLabels = TRUE,
                biplot = TRUE,
                extVariables = medians,
                arrowThreshold = 0.9) 
```

Instead of having one bi-plot related to a specific type of statistics, for 
example channel medians, one can try to associate the axes to different types 
of channel statistics at once. In the next plot, we represent such bi-plots 
for channel medians, 25% and 75% quantiles, and standard deviations. 

The 'faceting-alike' plot is obtained thanks to the 
`ggplotSampleMDSWrapBiplots()` function, which internally calls 
`ggplotSampleMDS()` function several times, and arrange the obtained outputs 
on a single plot.   

```{r plotMDS1_stats}
statFUNs = c("median" = stats::median,
             "Q25" = function(x, na.rm) {
                 stats::quantile(x, probs = 0.25)
             },
             "Q75" = function(x, na.rm) {
                 stats::quantile(x, probs = 0.75)
             },
             "standard deviation" = stats::sd)
chStats <- channelSummaryStats(OMIP021Sim1, statFUNs = statFUNs)
```

```{r plotMDS1_biplot_facetting, fig.align='center', fig.asp = 1, fig.wide = TRUE}
ggplotSampleMDSWrapBiplots(
    mdsObj1, 
    extVariableList = chStats,
    ncol = 2,
    pData = pData1,
    pDataForColour = "original_sample",
    pDataForShape = "original_sample",
    displayPointLabels = FALSE,
    displayArrowLabels = TRUE,
    repelArrowLabels = TRUE,
    arrowThreshold = 0.9,
    displayLegend = FALSE) 
```

Note that the last plot, with arrows corresponding to 
channel standard deviations, ganerates a warning indicating that the arrow 
labels could not be displayed. This is due to the fact that too many channel 
standard deviations are strongly negatively correlated with the x axis. 
When encountering such visual issues, it is advised to generate series of 
bi-plots, with subsets of channel statistics, in order to better identify 
the strongly correlated ones. One example is provided below:

```{r plotMDS1_biplot_stddev, fig.align='center', fig.asp = 2.5, fig.wide = TRUE}
stdDevs <- list()
stdDevs[["std dev of channels 1 to 6"]] <- 
    chStats[["standard deviation"]][,1:6]
stdDevs[["std dev of channels 7 to 12"]] <- 
            chStats[["standard deviation"]][,7:12]
stdDevs[["std dev of channels 13 to 18"]] <- 
            chStats[["standard deviation"]][,13:18]
ggplotSampleMDSWrapBiplots(
    mdsObj1, 
    ncol = 1,
    extVariableList = stdDevs,
    pData = pData1, 
    pDataForColour = "original_sample",
    pDataForShape = "original_sample",
    displayPointLabels = FALSE,
    displayArrowLabels = TRUE,
    repelArrowLabels = TRUE,
    arrowThreshold = 0.9) 
```

# Handling heavy datasets

Computing Earth Mover's Distances between all sample pairs of large data sets 
(e.g. with hundreds of samples), is a heavy computational task.  

First, loading the whole data set as a `flowCore::flowSet()` in RAM at once, 
might not be possible due to its size. Second, calculating a matrix of 
pairwise distances, has a computational complexity of O(N2), which can lead to 
very long computation times for large data sets.  

Therefore, the `CytoMDS` package provides several mechanisms allowing to
mitigate these issues.

## Loading flow frames dynamically during distance matrix computation

In order to be able to handle data set of greater size than the available 
computer RAM, the `pairwiseEMDDist()` function allows for differential input 
mode, where:
- the input samples are NOT provided directly via a `flowCore::flowSet`, but
- the user provides the nb of samples, and a user-written flow frame loading 
function that will be called to dynamically load the *i*th sample upon request, 
and optionally additional arguments.

Typically, the flow frame loading function provided by the user shall describe 
how to read the *i*th sample from disk. 

In the below, an example using data set `OMIP021Sim1` is provided. This is  
for illustrative purpose only, as this data set is light enough to reside fully 
in memory. In such a workflow, it is advised, as a preliminary step, to store 
all scale transformed samples on disk. Here we do this in a temporary directory. 
Note this scale transformation could also be done on the fly, as part 
of the flow frame loading function. However, this would then require 
scale transforming the channel intensities a number of times 
during distance matrix calculation, which would be inefficient.

```{r temporary_store}
storageLocation <- suppressMessages(base::tempdir())

nSample <- length(OMIP021Sim1)
fileNames <- file.path(
    storageLocation,
    paste0("OMIP021Sim1_Sample", seq_len(nSample), ".rds"))

for (i in seq_len(nSample)) {
    saveRDS(OMIP021Sim1[[i]], 
            file = fileNames[i])
}
```

Then, we now call the `pairwiseEMDDist()` function, and specify a loading 
function.

```{r distance_calc_loading_otf}
pwDist1Again <- pairwiseEMDDist(
    x = nSample,
    loadFlowFrameFUN = function(ffIndex, theFiles){
        readRDS(file = theFiles[ffIndex])
    },
    loadFlowFrameFUNArgs = list(theFiles = fileNames),
    verbose = TRUE
)
```

## Using BiocParallel to parallelize distance matrix computation

Finally, `CytoMDS` pairwise distance calculation supports parallelization of 
distance matrix computation, through the use of `BiocParallel` package.  

When parallelization is used, the calculation engine will automatically 
create worker tasks corresponding to specific blocks of the distance matrix 
to be calculated.  

Here below is an example, using `BiocParallel::SnowParam()` backbone. 

```{r distance_calc_loading_otf_parallel}
bp <- BiocParallel::SnowParam(
    stop.on.error = FALSE,
    progressbar = TRUE)
pwDist1Last <- suppressWarnings(pairwiseEMDDist(
    x = nSample,
    loadFlowFrameFUN = function(ffIndex, theFiles){
        readRDS(file = theFiles[ffIndex])
    },
    loadFlowFrameFUNArgs = list(theFiles = fileNames),
    useBiocParallel = TRUE,
    BPPARAM = bp))
```
The obtained distances - as displayed in the below histogram - are exactly 
the same as before.

```{r dist1ShowHistAgain, fig.align='center', fig.wide = TRUE}
distVec1Last <- pwDist1Last[upper.tri(pwDist1Last)]
distVecDF1Last <- data.frame(dist = distVec1Last)
pHist1Last <- ggplot(distVecDF1Last, mapping = aes(x=dist)) + 
    geom_histogram(fill = "darkgrey", col = "black", bins = 15) + 
    theme_bw() + ggtitle("EMD distances for data set 1 - parallel computation")
pHist1Last
```


# Session information {-}

```{r sessioninfo, echo=FALSE}
sessionInfo()
```


# References {-}
